---
title: "Rtglm: package overview"
date: "`r Sys.Date()`"
output:
   rmarkdown::html_vignette:
     toc: true
     toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

*Rtglm* provides a way to

1. Reproduce results of *EpiEstim*
1. Rely on GAM for temporal smoothing of Rt
1. Rely on GAM for spatial smoothing of Rt


```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# Installing the package

To install the current stable, CRAN version of the package, type:

To benefit from the latest features and bug fixes, install the development, *github* version of the package using:
```{r install2, eval=FALSE}
devtools::install_github("pnouvellet/Rtglm",build = TRUE)
devtools::install_github("pnouvellet/MCMCEpiEstim",build = TRUE)
```

Note that this requires the package *devtools* installed.
We then instal few additional useful package for this vignette.

```{r}
rm(list=ls())
library(Hmisc)
library(EpiEstim)
library(incidence)
library(projections)
library(ggplot2)
library(Rtglm)
library(MCMCEpiEstim)
```


# Reproducing EpiEstim results in a Rt.glm framework

First, we load data (from EpiEstim), e.g. the daily incidence of the 1918 influenza in Baltimore.

```{r}

# load data
data <- data(Flu1918)

# make a column for local and impated cases in incidence
Flu1918$incidence <- data.frame(local = Flu1918$incidence,
                                imported = 0)
Flu1918$incidence$imported[1] <- Flu1918$incidence$local[1]
Flu1918$incidence$local[1] <- 0
    
# plot incidence
plot(rowSums(Flu1918$incidence),
     main = 'Flu1918', bty = 'n', pch = 16,
     ylab = 'incidence', xlab = 'time')


```


## Running EpiEstim 

### set the Prior distribution for Rt

```{r}

# prior: exponential distribution with mean 1
mean_prior = 5
std_prior = 5
para_prior <- epitrix::gamma_mucv2shapescale(mu = mean_prior,cv = std_prior/mean_prior)
hist(rgamma(n = 1e3,shape = para_prior$shape, scale = para_prior$scale),
     main = 'prior dist. Rt', bty = 'n', 
     ylab = 'dist', xlab = 'Rt' )

```


### Run EpiEstim

Here we run EpiEstim using a wrapper to obtain standardized output. the wrapper is part of Rt.glm.
We run here EpiEstim with a 7 day time-window, with and without overlapping time-windows.

```{r,Rt}

# time windows
t_window <- c(7,7)
overlap <- c(TRUE,FALSE)

# output
Rt_EpiEstim <- list()

for(i in 1:length(t_window)){
  
  res <- Rtglm::EpiEstim_wrap(I_incid = Flu1918$incidence, 
                              si_distr = Flu1918$si_distr, 
                              t_window = t_window[i], 
                              overlap = overlap[i],
                              mean_prior = mean_prior, std_prior = std_prior)
  
  Rt_EpiEstim[[i]] <- res
}


```


## Run the same analyses with a glm

```{r}


# output
Rt_glm <- list()

for(i in 1:length(t_window)){
  res <- Rtglm::glm_Rt_wrap(I_incid = Flu1918$incidence, 
                            si_distr = Flu1918$si_distr, 
                            t_window = t_window[i], 
                            overlap = overlap[i])
  
  
  Rt_glm[[i]] <- res
}



```


## Compare glm and EpiEstim

```{r,fig.height = 6, fig.width = 8}

layout(matrix(c(1,1,1,1,2,2,2,2),nrow = 4, ncol = 2, byrow = TRUE))

for(i in 1:length(t_window)){
  # plot incidence
  plot(rowSums(Flu1918$incidence),
       main = 'Flu1918', bty = 'n', pch = 16,
       ylab = 'incidence', xlab = 'time')
  
  # plot EpiEstim and Rt.glm estimates
  a <- Rt_EpiEstim[[i]]$Rt[,c('t','Mean','low_Quantile','high_Quantile')]
  b <- Rt_glm[[i]]$Rt[,c('t','Mean','low_Quantile','high_Quantile')]
  
  res <- plot_compare2Rt(a = a, b = b,
                         t_window = t_window[i],
                         overlap = overlap[i],
                         Corr=TRUE)
}

```

# Temporal smoothing

```{r}

si <- Flu1918$si_distr
I0 <- incidence::as.incidence(x = 30, dates = 1, interval = 1)

```


## simulate an Rt

```{r}
x <- 1:101
B <- 30
A <- .2
Rt <- data.frame(t = x, 
                  Rt = A*sin((x+4)*2*pi/B)+1)

plot(Rt$t,Rt$Rt,type='l',lwd = 3,
     bty = 'n', xlab = 'Rt', ylab = 'time', ylim = c(0,1.5),
      xaxt='n',  yaxt='n', cex.lab=1.5)
axis(side = 1, at = c(0,50,100))
axis(side = 2, at = c(0,.50,1, 1.5))
abline(h = 1,col = 'red3', lty = 2)
```

## Simulate incidence

In simulating incidence, we make use of code available in *MCMCEpiEstim* R-package.

```{r}
set.seed(1)
res <- MCMCEpiEstim::project_fct(I0 = I0,
                                 Rt = Rt,
                                 n_loc = 1,
                                 t_max = nrow(Rt),
                                 si = si,
                                 p = 1,
                                 model = 'poisson')

sim_incidence <- res$I_true
plot(sim_incidence$`sim 1`,
     main = 'simulated incidence', bty = 'n', pch = 16,
     ylab = 'incidence', xlab = 'time')

# format incidence
data <- data.frame(local = c(0,sim_incidence[2:nrow(sim_incidence),2]),
                   imported = c(sim_incidence[1,2],rep(0,nrow(sim_incidence)-1) ))
```




## EpiEstim Estimates

```{r}
mean_prior = 5
std_prior = 5

# time windows
t_window <- 7 
overlap <- TRUE

# output
Rt_EpiEstim <- Rtglm::EpiEstim_wrap(I_incid = data,
                                    si_distr = si,
                                    t_window = t_window,
                                    overlap = overlap,
                                    mean_prior = mean_prior, std_prior = std_prior)
```


## Rt gam

```{r}
# output
Rt_gam <- Rtglm::gam_Rt_wrap(I_incid = data,
                             si_distr = si)
```


## check simulations

plot of incidence/Rts vs. time 

```{r,fig.height = 6, fig.width = 8}
layout(matrix(c(1,1,1,1,2,2,2,2),nrow = 4, ncol = 2, byrow = TRUE))

# plot incidence
plot(rowSums(data),
     main = 'simulated incidence', bty = 'n', pch = 16,
     ylab = 'incidence', xlab = 'time')

# plot simulated Rt
ylim <- c(0,1.5)
plot(Rt$t, Rt$Rt,type = 'l', lwd=2, 
     ylim = ylim,bty = 'n',
     xlab = 'time',ylab = 'Rt',
     main = '')

# plot EpiEstim estimates
f <- which( !is.na(Rt_EpiEstim$Rt$Mean) )
lines(Rt_EpiEstim$Rt$t[f], Rt_EpiEstim$Rt$Mean[f],col='red3')
polygon(c(Rt_EpiEstim$Rt$t[f],rev(Rt_EpiEstim$Rt$t[f])),
        c(Rt_EpiEstim$Rt$low_Quantile[f],
          rev(Rt_EpiEstim$Rt$high_Quantile[f])), 
        col = rgb(1,0,0,.1), border = NA)

# plot Rt.glm estimates
f <- which( !is.na(Rt_gam$Rt$Mean) )
lines(Rt_gam$Rt$t[f], Rt_gam$Rt$Mean[f],col='green4')
polygon(c(Rt_gam$Rt$t[f],rev(Rt_gam$Rt$t[f])),
        c(Rt_gam$Rt$low_Quantile[f],
          rev(Rt_gam$Rt$high_Quantile[f])), 
        col = rgb(0,1,0,.2), border = NA)

legend('bottomleft',legend = c('simulated','EpiEsitm','Rt.gam'), 
       lwd=1, col = c('black','red3','green4'), bty = 'n')
  

```


# Spatial smoothing

## Simulate Rt

## Simulate incidence

## EpiEstim Estimates

## Rt.glm estimate

## Compare estimates