---
title: "Rtglm: package overview"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: yes
# output:
#    rmarkdown::html_vignette:
#      toc: true
#      toc_depth: 2
# vignette: >
#   %\VignetteIndexEntry{Overview}
#   %\VignetteEngine{knitr::rmarkdown}
#   %\VignetteEncoding{UTF-8}
---

*Rtglm* provides a way to

1. Reproduce results of *EpiEstim*
1. Rely on GAM for temporal smoothing of Rt
1. Rely on GAM for spatial smoothing of Rt


```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# Installing the package

To install the current stable, CRAN version of the package, type:

To benefit from the latest features and bug fixes, install the development, *github* version of the package using:
```{r install, eval=FALSE}
devtools::install_github("pnouvellet/Rtglm",build = TRUE)
devtools::install_github("pnouvellet/MCMCEpiEstim",build = TRUE)
```

Note that this requires the package *devtools* installed.
We then install few additional useful package for this vignette.

```{r, library}
rm(list=ls())
library(Hmisc)
library(EpiEstim)
library(incidence)
library(projections)
library(ggplot2)
library(Rtglm)
library(MCMCEpiEstim)
library(sf)
```


# Reproducing EpiEstim results in a Rt.glm framework


First, we load data (from EpiEstim), e.g. the daily incidence of the 1918 influenza in Baltimore.


```{r,incidence_Flu}

# load data
data <- data(Flu1918)

# make a column for local and imported cases in incidence
Flu1918$incidence <- data.frame(local = Flu1918$incidence,
                                imported = 0)
# assume first case is imported and others are local cases.
Flu1918$incidence$imported[1] <- Flu1918$incidence$local[1]
Flu1918$incidence$local[1] <- 0
    
# plot incidence
plot(rowSums(Flu1918$incidence),
     main = 'Flu 1918', bty = 'n', pch = 16,
     ylab = 'incidence', xlab = 'time')


```






## EpiEstim estimate



First, we set the prior distribution for Rt

```{r,setupRt}

# prior distribution for Rt, gamma distribution.
mean_prior = 5
std_prior = 5
para_prior <- epitrix::gamma_mucv2shapescale(mu = mean_prior,cv = std_prior/mean_prior)
hist(rgamma(n = 1e3,shape = para_prior$shape, scale = para_prior$scale),
     main = 'prior dist. Rt', bty = 'n', 
     ylab = 'dist', xlab = 'Rt' )

```

Here we run EpiEstim using an Rtglm wrapper to obtain standardized output. the wrapper is part of Rtglm.
We run EpiEstim with a 7 day time-window, with and without overlapping time-windows.

```{r,Rt_EpiEstim_flu}

# time windows
t_window <- c(7,7) # 7 days time-windows
overlap <- c(TRUE,FALSE) # overlapping or non-overlapping time-windows

# output
Rt_EpiEstim <- list()

for(i in 1:length(t_window)){ # estimate Rt for all time-windows above
  
  res <- Rtglm::EpiEstim_wrap(I_incid = Flu1918$incidence, 
                              si_distr = Flu1918$si_distr, 
                              t_window = t_window[i], 
                              overlap = overlap[i],
                              mean_prior = mean_prior, std_prior = std_prior)
  
  Rt_EpiEstim[[i]] <- res
}


```


## Rt.glm estimate

```{r, Rt_glm_flu}


# output
Rt_glm <- list()

# same analyses but using the GLM formulation
for(i in 1:length(t_window)){
  res <- Rtglm::glm_Rt_wrap(I_incid = Flu1918$incidence, 
                            si_distr = Flu1918$si_distr, 
                            t_window = t_window[i], 
                            overlap = overlap[i])
  
  
  Rt_glm[[i]] <- res
}



```


## Compare estimates

Plot the resulting Rt estimates.

```{r,flu_fig,fig.height = 6, fig.width = 8}

layout(matrix(c(1,1,1,1,2,2,2,2),nrow = 4, ncol = 2, byrow = TRUE))

for(i in 1:length(t_window)){
  # plot incidence
  plot(rowSums(Flu1918$incidence),
       main = 'Flu1918', bty = 'n', pch = 16,
       ylab = 'incidence', xlab = 'time')
  
  # plot EpiEstim and Rt.glm estimates
  a <- Rt_EpiEstim[[i]]$Rt[,c('t','Mean','low_Quantile','high_Quantile')]
  b <- Rt_glm[[i]]$Rt[,c('t','Mean','low_Quantile','high_Quantile')]
  
  res <- plot_compare2Rt(a = a, b = b,
                         t_window = t_window[i],
                         overlap = overlap[i],
                         Corr=TRUE)
}

```

# Estimating Rt with a temporal smoothing


First, we characterising the serial interval.

```{r, temporal_setup}

si <- Flu1918$si_distr
I0 <- incidence::as.incidence(x = 30, dates = 1, interval = 1)

```


Then, we simulate an Rt trend over time, here we choose a sinusoidal pattern with a period of 30days and amplitude of 0.2 centered on 1.

```{r,Rt_temporal}
x <- 1:101
B <- 30
A <- .2
Rt <- data.frame(t = x, 
                  Rt = A*sin((x+4)*2*pi/B)+1)

plot(Rt$t,Rt$Rt,type='l',lwd = 3,
     bty = 'n', xlab = 'Rt', ylab = 'time', ylim = c(0,1.5),
      xaxt='n',  yaxt='n', cex.lab=1.5)
axis(side = 1, at = c(0,50,100))
axis(side = 2, at = c(0,.50,1, 1.5))
abline(h = 1,col = 'red3', lty = 2)
```

## Simulate incidence

In simulating incidence, we make use of code available in *MCMCEpiEstim* R-package.

```{r, incidence_temporal}
set.seed(1)
# make simulated incidence for a given initial conditions (I0), an Rt pattern (Rt), 
# for a single location (n_loc = 1), using a Poisson offspring distribution (model), 
# and assuming all cases are reported (below: p = 1).
res <- MCMCEpiEstim::project_fct(I0 = I0,
                                 Rt = Rt,
                                 n_loc = 1,
                                 t_max = nrow(Rt),
                                 si = si,
                                 p = 1,
                                 model = 'poisson')

sim_incidence <- res$I_true
# visualise the simulated incidence
plot(sim_incidence[,2],
     main = 'simulated incidence', bty = 'n', pch = 16,
     ylab = 'incidence', xlab = 'time')

# format incidence with local and imported cases as before.
data <- data.frame(local = c(0,sim_incidence[2:nrow(sim_incidence),2]),
                   imported = c(sim_incidence[1,2],rep(0,nrow(sim_incidence)-1) ))
```




## Estimate Rt using EpiEstim

```{r, Rt_EpiEstim_temporal}
mean_prior = 5
std_prior = 5

# time windows
t_window <- 7 
overlap <- TRUE

# output
Rt_EpiEstim <- Rtglm::EpiEstim_wrap(I_incid = data,
                                    si_distr = si,
                                    t_window = t_window,
                                    overlap = overlap,
                                    mean_prior = mean_prior, std_prior = std_prior)
```


## Estimate Rt using Rtglm

```{r,Rt_glm_temporal}
# output
Rt_gam <- Rtglm::gam_Rt_wrap(I_incid = data,
                             si_distr = si)
```


## Compare estimates

Visualise the estimated Rt trends estimated with EpiEstim, and with Rtglm using a temporal smooth. 

```{r, temporal_figures, fig.height = 6, fig.width = 8}
layout(matrix(c(1,1,1,1,2,2,2,2),nrow = 4, ncol = 2, byrow = TRUE))

# plot incidence
plot(rowSums(data),
     main = 'simulated incidence', bty = 'n', pch = 16,
     ylab = 'incidence', xlab = 'time')

# plot simulated Rt
ylim <- c(0,1.5)
plot(Rt$t, Rt$Rt,type = 'l', lwd=2, 
     ylim = ylim,bty = 'n',
     xlab = 'time',ylab = 'Rt',
     main = '')

# plot EpiEstim estimates
f <- which( !is.na(Rt_EpiEstim$Rt$Mean) ) # remove time-points where Rt cannot be estimated
lines(Rt_EpiEstim$Rt$t[f], Rt_EpiEstim$Rt$Mean[f],col='red3')
polygon(c(Rt_EpiEstim$Rt$t[f],rev(Rt_EpiEstim$Rt$t[f])),
        c(Rt_EpiEstim$Rt$low_Quantile[f],
          rev(Rt_EpiEstim$Rt$high_Quantile[f])), 
        col = rgb(1,0,0,.1), border = NA)

# plot Rt.glm estimates
f <- which( !is.na(Rt_gam$Rt$Mean) ) # remove time-points where Rt cannot be estimated
lines(Rt_gam$Rt$t[f], Rt_gam$Rt$Mean[f],col='green4')
polygon(c(Rt_gam$Rt$t[f],rev(Rt_gam$Rt$t[f])),
        c(Rt_gam$Rt$low_Quantile[f],
          rev(Rt_gam$Rt$high_Quantile[f])), 
        col = rgb(0,1,0,.2), border = NA)

legend('bottomleft',legend = c('simulated','EpiEsitm','Rt.gam'), 
       lwd=1, col = c('black','red3','green4'), bty = 'n')
  

```


# Spatial smoothing

We first load a map of the Unitary Authorities of the United Kingdom.

```{r, map_uk}
if (file.exists('ukMap.rds')){
  uk <- readRDS('ukMap.rds')
}else{
  # Define the raw URL
  rds_url <- "https://raw.githubusercontent.com/pnouvellet/Rtglm/master/vignettes/ukMap.rds"
  # Temporary file path
  temp_file <- tempfile(fileext = ".rds")
  # Download the file
  download.file(rds_url, temp_file, mode = "wb")
  # Read the RDS file
  uk <- readRDS(temp_file)
}
class(uk)
```


## Simulate Rt trends over space

We use the same method as in:
Rtglm: Unifying estimation of the time-varying reproduction number, Rt , under the Generalised Linear and Additive Models.

In Summary, We construct a kernel formed of 3 distinct 2-dimension Normal distributions, with means chosen randomly among centroid of the UA of the UK and with variance randomly chosen between 0.5 and 2.

```{r, Rt_spatial}
library(mvtnorm)
set.seed(1)

# range of simulated Rt
Rt_min_max <- data.frame(max = 2,
                        min = 0.5)

# range of the variance of the bi-variate Normals
var_range <- c(.5,2)

# locations of 3 peak in Rt and associated variances
r_loc <- sample(x = 1:nrow(uk),size = 3,replace = TRUE)
Mus <- list(c(uk$cent.x[r_loc[1]],uk$cent.y[r_loc[1]]),
            c(uk$cent.x[r_loc[2]],uk$cent.y[r_loc[2]]),
            c(uk$cent.x[r_loc[3]],uk$cent.y[r_loc[3]]))
Sigmas <- runif(n = 3, min = var_range[1], max = var_range[2])

# simulate Rt
R_true <- data.frame(GID_2 = uk$GID_2,
                     cent.x = uk$cent.x, cent.y = uk$cent.y,
                     Rt = NA)

# simulate Rt
R_true$Rt <- Rtglm::kernel_rt(map = uk,
                       mu = Mus, sigma = Sigmas,
                       Rt_min_max = Rt_min_max)
```

Optional plot to visualise the true Rt

```{r, Rt_spatial_plot, eval = FALSE}


temp <- merge(uk,R_true)

ggplot(temp, aes(fill = Rt)) +
  ggplot2::geom_sf() +
  scale_fill_viridis_c()


```


## Simulate incidence

We first define some useful function to back-calculate incidence expected given Rt and a total incidence.

```{r, useful}
# function that recovers growth rate (r) from Reproduction number (R) and serial interval
# taken from Walling and Lipsitch 2006
r_2_R <- function(r){
  R <- 1/sum(exp(-r*0:(length(si)-1))*si)
  return(R)
}

# r_2_R(r = -0.5)
# function to numerically find the root of equation above, e.g. find r given R and the serial interval
findRoot <- function(r,R){
  R_check <- r_2_R(r)
  return( (R_check - R)^2 )
}
# test
# r <- optim(par = 0, fn = findRoot, method = 'BFGS', hessian=TRUE, R = 2 )$par
```

simulate incidence

```{r, incidence_spatial}
ini_I <- I_sim <- list()
I0 <- 30
t_sim <- 5 # simulate 5 days of incidence to be used in the inference

# initialise incidence according to Rt, prior to the period of inference.
for (i in 1:nrow(uk)){   # do it for each locations
  R <- R_true$Rt[i] # location specific Rt
  # find growth rate associated with Rt above
  ini_r <- optim(par = 0, fn = findRoot,
                 method = 'BFGS',
                 hessian=TRUE, R = R )$par
  # we simulate 10 days incidence prior to inference period
  init_t <- 10
  # find the exponential growth/decline associated with Rt starting from 1 case
  temp <- exp(ini_r*1:init_t)
  # compute the associated overall infectivity
  temp2 <- EpiEstim::overall_infectivity(incid = temp,
                                         si_distr =  si)
  # adjust incidence such that if R=1, given the simulated incidence,
  # we would expect 30 cases on the first simulation day
  # on day 1 of simulation: E{I_11} = R* Sum{I{1:10}*si{10:1}}
  temp3 <- temp/temp2[init_t]*I0 
  # return integer incidence object
  ini_I[[i]] <-  incidence::as.incidence(x = round(temp3),
                                         dates = 1:init_t, interval = 1)
}

# now simulate some stochastic incidence given the above
for (i in 1:nrow(uk)){
  R <- R_true$Rt[i]
  # simulate incidence
  I <- as.data.frame(projections::project(x = ini_I[[i]],
                                          R = R,
                                          n_sim = 1, # number of simulation
                                          si = si[-1],
                                          n_days = t_sim,
                                          instantaneous_R = TRUE))
  
  
  d_incidence <- data.frame(t = c(1:ini_I[[i]]$timespan,I[,1]),
                            incidence = c(ini_I[[i]]$counts,I[,2]))
  
  # specify importation (e.g. the initialisation)
  I_corr <- data.frame(local = d_incidence$incidence,
                       imported = 0)
  # correct initial case as imported
  I_corr$imported[1:init_t] <- d_incidence$incidence[1:init_t]
  I_corr$local[1:init_t] <- 0
  
  
  I_sim[[i]] <- I_corr
}
```

## Estimates using EpiEstim 


### By UA

We independently estimate one Rt for each UA.

```{r, Rt_EpiEstim_spatial_UA}
res <-  data.frame(matrix(NA, nrow = nrow(uk), ncol = 5))
names(res) <-  c('Mean','Std',
                 'low_Quantile','Median','high_Quantile')

for (i in 1:nrow(uk)){
  
  res_EE <- Rtglm::EpiEstim_sp_wrap(I_incid = I_sim[[i]],
                                    si_distr = si,
                                    t_ini = init_t)
  
  
  res[i,] <- c(res_EE$R$Mean, res_EE$R$Std,
               res_EE$R$low_Quantile, res_EE$R$Median, res_EE$R$high_Quantile)
}

names(res) <- paste0('UA_',names(res))
R_UA <- res
```

### By mid-grid

We compute the aggregated incidence within a grid comprising multiple UA. Then, as above, we independently estimate one Rt for each grid.

```{r, Rt_EpiEstim_spatial_grid}
res <-  data.frame(matrix(NA, nrow = nrow(uk), ncol = 5))
names(res) <-  c('Mean','Std',
                 'low_Quantile','Median','high_Quantile')

unique_grid <- unique(uk$grid2)

for (i in 1:length(unique_grid)){
  # find which UA belong to the grid
  f <- which(uk$grid2 %in% unique_grid[i])
  # initialise incidence with the first UA belonging to the grid
  I_corr <- I_sim[[ f[1] ]]
  # add incidence from other UAs belong to the grid
  if(length(f)>1){
    for(l in 2:length(f)){
      I_corr <- I_corr+I_sim[[ f[l] ]]
    }
  }
  
  # estimate Rt for the grid
  resEE <- Rtglm::EpiEstim_sp_wrap(I_incid = I_corr,
                                   si_distr = si,
                                   t_ini = init_t)
  
  # save output
  res[f,] <- matrix(data = c(resEE$R$Mean, resEE$R$Std,
                             resEE$R$low_Quantile, resEE$R$Median,
                             resEE$R$high_Quantile), nrow = length(f), ncol = 5,
                    byrow = TRUE)
  
}

names(res) <- paste0('grid_',names(res))
R_grid <- res
```


## Rt.glm estimate

Estimate Rt across space, using a smooth spatial function for Rt.

```{r, Rt_glm_spatial}
res_gam <- Rtglm::gam_Rt_sp_wrap(I_incid = I_sim,
                          si_distr = si,
                          x=uk$cent.x,
                          y=uk$cent.y)

res <- res_gam$Rt[,-1]
names(res) <- paste0('gam_',names(res))
R_gam <- res
```

## Compare estimates

Visualise the results

## maps of Rts

First, we prepare the dataframe for ggplot.

```{r, spatial_figures_prep, eval = TRUE}
uk <- st_sf(cbind(uk,
              R_UA,
              R_grid,
              R_gam))

temp <- R_true[,c(1,4)]
names(temp)[2] <- 'R'

uk <- merge(uk,temp)
```

We plot the mean Rt estimates, using either EpiEstim at UA scale, at grid scale, and using Rtglm with a spatial smooth.

```{r,spatial_figures_mean}
ggplot(uk, aes(fill = R)) +
  ggplot2::geom_sf() +
  ggtitle('Simulated Rt')+
  scale_fill_viridis_c(limits = c(0, 3),name = 'R',option = 'H') 

ggplot(uk, aes(fill = UA_Mean)) +
  ggplot2::geom_sf() +
  ggtitle('EpiEstim Rt - UA')+
  scale_fill_viridis_c(limits = c(0, 3),name = 'R',option = 'H') 


limits_x <- seq(min(uk$cent.x),max(uk$cent.x),length.out = 5)
limits_y <- seq(min(uk$cent.y),max(uk$cent.y),length.out = 5)

ggplot(uk, aes(fill = grid_Mean)) +
  ggplot2::geom_sf() +
  ggtitle('EpiEstim Rt - medium grid') +
  scale_fill_viridis_c(limits = c(0, 3),name = 'R',option = 'H') +
  geom_vline(xintercept=limits_x[-c(1,length(limits_x))], linetype="solid",col='red3')+
  geom_hline(yintercept=limits_y[-c(1,length(limits_x))], linetype="solid",col='red3') 

ggplot(uk, aes(fill = gam_Mean)) +
  ggplot2::geom_sf() +
  ggtitle('GAM Rt') +
  scale_fill_viridis_c(limits = c(0, 3),name = 'R',option = 'H')  
```


## plot uncertainties maps (CV)

We plot the mean Coefficient of Variation estimates, using either EpiEstim at UA scale, at grid scale, and using Rtglm with a spatial smooth.

```{r, spatial_figures_CV, eval = TRUE}
uk$UA_cvL <- log10(uk$UA_Std/uk$UA_Mean)
uk$grid_cvL <- log10(uk$grid_Std/uk$grid_Mean)
uk$gam_cvL <- log10(uk$gam_Std/uk$gam_Mean)

# range(c(temp$UA_cvL ,
#         temp$grid_cvL ,
#         temp$gam_cvL))

ggplot(uk, aes(fill = UA_cvL)) +
  ggplot2::geom_sf() +
  ggtitle('CV of EpiEstim Rt - UA')+
  scale_fill_viridis_c(limits = c(-2.1,0),name = 'Cv (R)',option = 'H',
                       breaks = c(-3,-2,-1,0), labels = 10^c(-3,-2,-1,0)) 

ggplot(uk, aes(fill = grid_cvL)) +
  ggplot2::geom_sf() +
  ggtitle('CV of EpiEstim Rt - medium grid') +
  scale_fill_viridis_c(limits = c(-2.1,0),name = 'Cv (R)',option = 'H',
                       breaks = c(-3,-2,-1,0), labels = 10^c(-3,-2,-1,0)) +
  geom_vline(xintercept=limits_x[-c(1,length(limits_x))], linetype="solid",col='red3')+
  geom_hline(yintercept=limits_y[-c(1,length(limits_x))], linetype="solid",col='red3') 

ggplot(uk, aes(fill = gam_cvL)) +
  ggplot2::geom_sf() +
  ggtitle('CV of Gam Rt') +
  scale_fill_viridis_c(limits = c(-2.1,0),name = 'Cv (R)',option = 'H',
                       breaks = c(-3,-2,-1,0), labels = 10^c(-3,-2,-1,0))  
```

